---
title: "Code report"
author: "Pham Xuan Huy Nguyen"
date: "2023-05-11"
output: html_document
---

## brackenPCA.R - ORIGINAL CODE + My comment

### Load packages + args

This part seems okay, nothing was weird to me at least.\
But there are some stuffs that worth mentioning:\
1. the *raw_counts* variable was never used again\
2. could it be the *raw_counts* is also the *raw_counts_matrix* in the next part?\
3. I cant not run the function *read_table* in my R Studio although I loaded all the required packages, alternatively I use *read.table* instead

```{r}
# 
library(dplyr)
library(tibble)
library(ggplot2)
library(DESeq2)

# instruct R to take information from the command line
args = commandArgs(trailingOnly=TRUE)

## test if there is at least one argument: if not, return an error
if (length(args)< 2) {
  stop("At least two arguments must be supplied: 1) tab-separated bracken raw count matrix (rows = samples, columns = taxa), 2) metadata file where sample IDs match count matrix. 3) Prefix (optional)", call.=FALSE)
} else if (length(args)<3) {
  # default output file
  args[3] = str_split_fixed(args[1], "\\.", n = 2)[1]
}


raw_counts <- read_table(args[1], col_names = T) 
metadata <- read_table(args[2], col_names = T)
prefix = args[3]

# if you want to run this in the Rstudio, you can read in your files directly
# raw_counts <- read_table("name_of_file")
```

### Prep data

Assuming the *bracken_est_reads* is the variable which holds the merged data.\
I guess the *\~replace_na(.x, 0)* is you gonna replace the NA value with 0.\
But I cant not run the code with the *replace...*, so I have to removed it and do manually.\
Additionally, I think it doesn't matter if I used the first code or not, as long as I can replace the NA with 0?

```{r}
## Run the pca
# you should use scaled, normalized data when running the pca
# all NAs are converted to 0. 
raw_counts_matrix <- bracken_est_reads %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>%
  column_to_rownames("sample_id") %>% 
  as.matrix()

# Because we cannot take the log of 0, we add 1 read
# this is called a shifted log transformation
```

This part is fine

```{r}
log_counts_matrix <- log2(raw_counts_matrix +1)
```

This is where the problem started to rise.\
So I guess, the *prcomp* is the one that perform the PCA, and by default the names of the principal components are named PC1 PC2 ect.\
But in *PCA.data.frame* we used the *PCA.scaled_log_counts* that was made from the *prcomp* and *left_join* it by the *sample_id* which was kinda messed up.\
Because on one side we have PC1 2..., on the *metadata* we have the original name P12... so joining the data made the data about hostplant or sth became NA.\

The *PCA.summary* was okay, I guess, but still the naming was PC1 2 ect, so the bar plot it made, the label of each bar was PC1 PC2 ect.\

```{r}
# calculate principal components
PCA.scaled_log_counts <- prcomp(log_counts_matrix, scale. = F, center = T)
PCA.summary <- as.data.frame(t((summary(PCA.scaled_log_counts))$importance)) %>% 
  rownames_to_column(var = "Principal Component")
# write_tsv(PCA.summary,paste0(prefix, "PCA.summary.tsv"))

# explore principal component axes
PCA.data.frame <- as.data.frame(PCA.scaled_log_counts$x) %>%
  rownames_to_column(var = "sample_id") %>% 
  left_join(metadata, by = "sample_id")
# write_tsv(PCA.data.frame,paste0(prefix, "PCA.samples.tsv"))
```

### Plotting

This one I guess I dont have any problem, but the problem was also mentioned (the bar plot label is PC1 2...)

```{r}
PCA.propVar <- ggplot(PCA.summary[1:6,], 
                      aes( x= `Principal Component`, y = `Proportion of Variance`)) +
  geom_bar(stat = "identity") +
  labs(x = "Principal component", "Prop. of variance") +
  geom_label(stat = "identity", aes(label = round(`Proportion of Variance`, 3))) +
  theme_classic(base_size = 15)
# ggsave(PCA.propVar, file =paste0(prefix, "PCA.plot.propVariance.pdf"), height = 4.5, width = 6)

```

Should *aes\_* be just *aes*? And here I guess the *aes(color = hostplant, shape= transect)* is not gonna work because of the problem I mentioned above, about them (hostplant, transect) having NA value.

```{r}
pca_plot <- list()
for (i in 2:5){
  PC_axes <- c(1,i)
  PC_columns <- c(paste0("PC",PC_axes[1]), paste0("PC",PC_axes[2]))
  PC_var <- c(paste0(PC_columns[1], " (", round(PCA.summary[PC_axes[1],3]*100, 1), "%)"), paste0(PC_columns[2], " (", round(PCA.summary[PC_axes[2],3]*100, 1), "%)"))
  
  pca_plot[[i-1]] <- ggplot(PCA.data.frame, aes_(x = as.name(PC_columns[1]), y = as.name(PC_columns[2]))) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    # geom_point(size = 2, stroke = 1.25)+
    geom_point(size = 2, stroke = 1.25, aes(color = hostplant, shape= transect))+
    scale_color_manual(values = c("#5E548E", "#32936F"), name = "Host Plant")+
    scale_shape_manual(values = c(1,2), name = "Transect") +
    labs(x = PC_var[1], y =PC_var[2]) +
    theme_minimal(base_size = 12) + theme(legend.position = "bottom")
  # ggsave(plot = pca.plot[[i-1]], paste0(prefix, "PCA.plot.",PC_columns[1], PC_columns[2],".pdf"), height = 5.5, width = 5)
}
pca_plot
```

POSSIBLE EXPLANATION: Could it be because the way I merge the data was not in the right way your code gonna read?\

## Can you try my way of merging the data and apply your code on it?
1. Clone the bracken subfolder in my github that contains the bracken data.
2. Follow the following code
```{r}
dir <- "" # Write the directory of the bracken folder inside the " "

files <- list.files(dir)

# Manually handle the first file
one <- read.table(files[1], header=TRUE, sep='\t') # Read file
name <- sub("\\..*", "", files[1]) # Change column name
colnames(one)[2] <- name

merge_data <- one # Set the first data frame as a variable for the loop
for (i in 2:length(files)){ # Loop through the files
  temp_data <- read.table(files[i], header=TRUE, sep='\t') # Read the df
  colnames(temp_data)[2] <- sub("\\..*", "", files[i]) # Change column name
  merge_data <- merge(merge_data, temp_data, all.x = TRUE, all.y = TRUE) # Merge
}

merge_data_trans <- t(merge_data) #Flip the data to the correct format for pca
```